---
title: "Ordination with vegan"
author: "Naupaka Zimmerman and Gavin Simpson"
date: "August 9, 2014 • ESA 2014"
bibliography: ordination.bib

output:
  ioslides_presentation:
    widescreen: true
    smaller: false
    keep_md: true
    highlight: tango
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "", cache = TRUE)
```

<!----------------------------slide boundary--------------------------------->

# Ordination plots

<!----------------------------slide boundary--------------------------------->

## Unconstrained ordination

**What does it mean?**

* First we look for major variation, then try to attribute it to environmental variation
* vs. constrained ordination, where we only want to see what can be explained by environmental variables of interest

<!----------------------------slide boundary--------------------------------->

## Unconstrained ordination

**What does it mean?**

* First we look for major variation, then try to attribute it to environmental variation
* vs. constrained ordination, where we only want to see what can be explained by environmental variables of interest

**Examples of unconstrained ordination:**

* Principal Components Analysis - PCA
* Nonmetric Multidimensional Scaling - NMDS

<!----------------------------slide boundary--------------------------------->

## Unconstrained ordination

**What does it mean?**

* First we look for major variation, then try to attribute it to environmental variation
* vs. constrained ordination, where we only want to see what can be explained by environmental variables of interest

**Examples of unconstrained ordination:**

* Principal Components Analysis - PCA
* **Nonmetric Multidimensional Scaling - NMDS**

<!----------------------------slide boundary--------------------------------->

## Before we get started

Housekeeping
```{r, eval=FALSE}
setwd("your/working/dir")
```

```{r, results='hide'}
library("vegan")
data(dune)
data(dune.env)
```
Data from : Jongman, R.H.G, ter Braak, C.J.F & van Tongeren, O.F.R. (1987). Data Analysis in Community and Landscape Ecology. Pudoc, Wageningen.

<!----------------------------slide boundary--------------------------------->

## Before we get started

```{r}
head(dune[,1:10], n=3)
```

<!----------------------------slide boundary--------------------------------->

## Before we get started

```{r}
head(dune.env, n=3)
summary(dune.env)
```

<!----------------------------slide boundary--------------------------------->

## Basic ordination and plotting

There are two more basic NMDS ordination functions:

* `isoMDS()` from the MASS package
* `monoMDS()` in vegan

Vegan also has a wrapper function for doing NMDS ordinations using best practices:

* `metaMDS()`

This will do handy things like try to standardize your data if necessary and perform rotation, among other things.

<!----------------------------slide boundary--------------------------------->

## Basic ordination and plotting

```{r, results='hide'}
dune.bray.ord <- metaMDS(dune, distance = "bray", k = 2, trymax = 50)
```

**Show in RStudio**

<!----------------------------slide boundary--------------------------------->

## Basic ordination and plotting (using all defaults)

```{r, fig.height=5}
plot(dune.bray.ord)
```


<!----------------------------slide boundary--------------------------------->

## Basic ordination and plotting (just plots)

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites")
```

<!----------------------------slide boundary--------------------------------->

## Basic ordination and plotting (just species)

```{r, fig.height=5}
plot(dune.bray.ord, display = "species")
```

<!----------------------------slide boundary--------------------------------->

## Site names instead of points

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", type = "t")
```

<!----------------------------slide boundary--------------------------------->

## Larger points

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", cex=2)
```

<!----------------------------slide boundary--------------------------------->

## Modifying the display of the points with environmental data 

* Color
* Shape
* Size

<!----------------------------slide boundary--------------------------------->

## Modifying the color of points {.smaller}
```{r, fig.height=4.5, strip.white=TRUE}
colors.vec <- c("red", "blue", "orange", "grey")
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord, display = "sites", cex=2, pch = 21, 
       col = colors.vec[dune.env$Management], 
       bg = colors.vec[dune.env$Management])
legend("topright", legend = levels(dune.env$Management), bty = "n",
                      col = colors.vec, pch = 21, pt.bg = colors.vec)
```

<!----------------------------slide boundary--------------------------------->

## Modifying the shape of points
```{r, fig.width=10, echo=FALSE}
par(mar = rep(0,4))
plot(c(-1, 26), 0:1, type = "n", axes = FALSE)
text(0:25, 0.4, 0:25, cex = 1.5)
points(0:25, rep(0.3, 26), pch = 0:25, bg = "grey", cex = 2)
```

<!----------------------------slide boundary--------------------------------->

## Modifying the shape of points {.smaller}
```{r, fig.height=3.5}
shapes.vec <- c(21, 22, 24)
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord, display = "sites", cex=2, bg = "black", 
       pch = shapes.vec[dune.env$Use])
legend("topright", legend = levels(dune.env$Use), bty = "n",
                      col = "black", pch = shapes.vec, pt.bg = "black")
```

<!----------------------------slide boundary--------------------------------->

## Modifying the shape and color of points {.smaller}
```{r, fig.height=3.5}
colors.vec <- c("red", "blue", "orange", "grey")
shapes.vec <- c(21, 22, 24)
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord, display = "sites", cex=2, 
       pch = shapes.vec[dune.env$Use], 
       col = colors.vec[dune.env$Management], 
       bg = colors.vec[dune.env$Management])
legend("topright", legend = levels(dune.env$Management), bty = "n",
                      col = colors.vec, pch = 21, pt.bg = colors.vec)
legend(1.5,1.05, legend = levels(dune.env$Use), bty = "n",
                      col = "black", pch = shapes.vec, pt.bg = "black")
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers

```{r, fig.height=5}
# Just points
plot(dune.bray.ord, display = "sites", cex=2)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", cex=2)
ordihull(dune.bray.ord,groups = dune.env$Management, label = TRUE)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", cex=2)
ordihull(dune.bray.ord,groups = dune.env$Management, label = TRUE, col = "blue")
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", cex=2)
ordihull(dune.bray.ord,groups = dune.env$Management, label = TRUE, col = "blue")
ordispider(dune.bray.ord,groups = dune.env$Management, label = TRUE)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers

```{r, fig.height=5}
# Plot first, then add layers
plot(dune.bray.ord, display = "sites", cex=2)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", cex=2)
ordispider(dune.bray.ord,groups = dune.env$Management, label = TRUE)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers - axes scaling

```{r, fig.height=5}
plot(dune.bray.ord, type = "n")
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers - axes scaling

```{r, fig.height=5}
plot(dune.bray.ord, type = "n")
points(dune.bray.ord,display = "sites", cex = 2)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers - axes scaling

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", type = "n")
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers - axes scaling

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord, display = "sites", cex = 2)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord,display = "sites", cex = 2)
ordispider(dune.bray.ord,groups = dune.env$Management, label = TRUE)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers

```{r, fig.height=5}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord, display = "sites", cex = 2)
ordiellipse(dune.bray.ord,groups = dune.env$Management, label = TRUE)
```

<!----------------------------slide boundary--------------------------------->

## Adding other layers
```{r, fig.height=5, message=FALSE}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord,display = "sites", cex = 2)
ordisurf(dune.bray.ord,dune.env$A1, add = TRUE)
```

<!----------------------------slide boundary--------------------------------->

## Vectors in ordination space

```{r}
dune.bray.ord.A1.fit <- envfit(dune.bray.ord,dune.env$A1, permutations = 1000)
dune.bray.ord.A1.fit
```

<!----------------------------slide boundary--------------------------------->

## Vectors in ordination space

```{r}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord,display = "sites", cex = 2)
plot(dune.bray.ord.A1.fit, add = TRUE)
ordisurf(dune.bray.ord,dune.env$A1, add = TRUE)
```


<!----------------------------slide boundary--------------------------------->

## References
